```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  fig.width = 6.3,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
knitr::opts_knit$set(root.dir = "..")
```

```{r libraries, include=FALSE}
library(tidyverse)
library(lubridate)
library(epiR)
source("utils/plotting.r")
source("utils/trans_risk.r")
source("utils/tex.r")
```

## Data

- Data collected at Masi clinic in 2019 (before the COVID-19 pandemic) and in 2021 (during pandemic with infection control measures in place).
- Data was collected from July to August in 2019 and from October to November in 2021.
- Bioaerosol sampling was performed twice per day, once in the morning before 11am and once in the afternoon before 2pm.
- The sampling durations were not the same, i.e. 2h in 2019 vs 3h in 2021. 
- To compare the results, we compute the Mtb concentration in the air as log DNA copies per muL per hour of sampling.
- Note that the exact sampling times were also not the same, which may affect the results.
- Furthermore, lab analysis was done by different people with possible influence on the results.
- Otherwise, collection device and technical protocols were the same.

```{r overview, include=FALSE}
df <- readRDS("data-clean/combined-data-wr.rds") %>%
  mutate(year = factor(year, levels = c(2019, 2021)))

clin <- read_csv("data-clean/clinical-data.csv") %>%
  mutate(year = factor(year, levels = c(2019, 2021)))

full_environ <- readRDS("data-clean/2019/environmental.rds") %>%
  rbind(readRDS("data-clean/2021/environmental.rds")) %>%
  filter(room == "Waiting room") %>%
  filter(date %in% df$date) %>%
  filter(between(hour(date_time), 8, 18)) %>%
  mutate(
    year = year(date),
    year = factor(year, levels = c(2019, 2021)),
    daytime = ifelse(hour(date_time) < 12, "Morning", "Afternoon"),
    daytime = factor(daytime, levels = c("Morning", "Afternoon"))
  )
```

```{r data-summary}
sprintf(
  "Total number of observations (study days): %i (%i)",
  nrow(df), n_distinct(df$date)
)

sprintf(
  "Number of observations: %i (2019) and %i (2021)",
  sum(df$year == 2019), sum(df$year == 2021)
)

sprintf(
  "Number of non-missing outcomes: %i (2019) and %i (2021)",
  sum(!is.na(df$dna_copies) & df$year == 2019),
  sum(!is.na(df$dna_copies) & df$year == 2021)
)
```

### Missingness

**Figure**: Missingness by type of data. 

```{r missingness, fig.height=10}
missing_df <- df %>%
  mutate(
    across(
      c(dna_copies, humidity),
      ~ is.na(.x)
    )
  ) %>%
  rowwise() %>%
  mutate(
    pt = ifelse(
      daytime == "Morning",
      any(is.na(c(pt_hr8, pt_hr9, pt_hr10, pt_hr11))),
      any(is.na(c(pt_hr12, pt_hr13, pt_hr14, pt_hr15)))
    )
  ) %>%
  ungroup()

missing_pl <- missing_df %>%
  pivot_longer(
    cols = c(dna_copies, humidity, pt),
    names_to = "variable",
    values_to = "missing"
  ) %>%
  mutate(
    variable = case_when(
      variable == "dna_copies" ~ "Outcome",
      variable == "humidity" ~ "Environmental",
      variable == "pt" ~ "Tracking"
    ),
    variable = factor(variable, levels = c(
      "Outcome",
      "Environmental",
      "Tracking"
    )),
    daytime = factor(daytime, levels = c("Morning", "Afternoon"))
  ) %>%
  arrange(date, daytime) %>%
  mutate(
    date_time = paste(
      format(date, format = "%b %d"),
      ifelse(daytime == "Morning", "(AM)", "(PM)")
    )
  ) %>%
  ggplot(aes(
    y = date_time,
    x = variable,
    fill = factor(missing, levels = c(TRUE, FALSE))
  )) +
  geom_tile() +
  facet_wrap(~year, scales = "free_y") +
  labs(
    x = NULL,
    fill = "Missing"
  ) +
  scale_fill_manual(values = c("grey", "white")) +
  theme_custom() +
  theme(
    legend.position = "bottom",
    axis.title = element_blank()
  )

save_plot(
  missing_pl,
  "results/missing-data.png",
  w = 14,
  h = 24
)
```

Interpretation:

- Outcome, environmental, and tracking data were not consistently available, especially in 2021.
- Thus, we can hardly relate the outcome to environmental conditions and room occupancy on a daily basis because often one variable is missing.
- Note that clinical data is always available and compared overall. 
- Frequency of TB patient visits is known for both years, but not what type of patients are visiting (HIV status, MDR, smear/GeneXpert grading).

&nbsp;

**Table**: Number of study days with complete measurements by type of data, only considering days with at least one variable non-missing.

```{r missingness-tbl}
missing_df %>%
  mutate(registered = is.na(registered)) %>%
  dplyr::select(year, date, registered, dna_copies, humidity, pt) %>%
  group_by(year, date) %>%
  summarise_all(function(x) all(!x)) %>%
  ungroup() %>%
  group_by(year) %>%
  summarise(across(c(-date), sum)) %>%
  set_names(c(
    "Year",
    "Clinical data",
    "Bio-aerosol sampling",
    "Environmental data",
    "Tracking data"
  )) %>%
  knitr::kable(align = "lrrrr")
```

&nbsp;

## Comparison of daily data

```{r custom-boxplot}
p_value_format <- function(p_value) {
  ifelse(
    p_value < .001, "p < 0.001",
    ifelse(p_value < 0.01, paste("p =", round(p_value, 3)),
      paste("p =", round(p_value, 2))
    )
  )
}

custom_boxplot <- function(dat) {
  p_value_1 <- ggpubr::compare_means(
    y ~ year,
    data = dplyr::filter(dat, daytime == "Morning"), method = "t.test"
  )$p.adj
  p_value_2 <- ggpubr::compare_means(
    y ~ year,
    data = dplyr::filter(dat, daytime == "Afternoon"), method = "t.test"
  )$p.adj
  p_value <- c(p_value_1, p_value_2)
  p_value <- p_value_format(p_value)
  dat %>%
    mutate(
      x2 = interaction(year, daytime),
      x2 = factor(x2, levels = c(
        "2019.Morning", "2021.Morning",
        "2019.Afternoon", "2021.Afternoon"
      ))
    ) %>%
    ggplot(aes(
      x = x2,
      y = y
    )) +
    geom_boxplot(aes(fill = year)) +
    geom_jitter(width = .1, height = 0, alpha = .5) +
    ggsignif::geom_signif(
      comparisons = list(c(1, 2), c(3, 4)),
      map_signif_level = FALSE,
      textsize = 8 / cm(1),
      annotations = p_value,
    ) +
    scale_fill_manual(
      values = wes_palette("Zissou1")[c(2, 4)]
    ) +
    scale_x_discrete(
      labels = c("Morning", "Morning", "Afternoon", "Afternoon"),
    ) +
    theme_custom() +
    theme(
      axis.title.x = element_blank(),
      legend.position = "bottom",
      legend.title = element_blank()
    )
}
```

```{r tracking}
pt_df <- df %>%
  dplyr::select(year, date, matches("pt_hr")) %>%
  pivot_longer(
    cols = -c(year, date),
    names_to = "daytime",
    values_to = "pt"
  ) %>%
  mutate(
    daytime = gsub("pt_hr", "", daytime),
    daytime = as.numeric(daytime),
    daytime = ifelse(daytime <= 11, "Morning", "Afternoon"),
    daytime = factor(daytime, levels = c("Morning", "Afternoon"))
  ) %>%
  group_by(year, date, daytime) %>%
  filter(all(!is.na(pt))) %>%
  summarise(
    y = sum(pt)
  ) %>%
  ungroup()

track_pl <- pt_df %>%
  custom_boxplot() +
  labs(
    y = "Person-time (hours)",
    title = "A | Person-time in clinic",
  ) +
  scale_y_continuous(
    limits = c(0, 350),
    expand = c(0, 0)
  )
```

```{r environ}
daily_max_co2 <- full_environ %>%
  filter(room == "Waiting room") %>%
  group_by(year, date, daytime) %>%
  summarise(co2 = max(co2, na.rm = TRUE)) %>%
  ungroup()

co2_pl <- daily_max_co2 %>%
  filter(!is.na(co2)) %>%
  rename(y = co2) %>%
  custom_boxplot() +
  labs(
    y = expression("Daily maximum CO"[2] * " (parts per million)"),
    title = expression(bold("C | CO"[2] * " concentration")),
  ) +
  scale_y_continuous(
    limits = c(400, 1600),
    labels = scales::comma
  )

aer_pl <- df %>%
  filter(!is.na(aer)) %>%
  rename(y = aer) %>%
  custom_boxplot() +
  labs(
    y = "Air changes per hour",
    title = "D | Air exchange rate",
  ) +
  scale_y_continuous(
    limits = c(0, 50),
    expand = c(0, 0)
  )
```

```{r tb-visits}
inc_pl <- df %>%
  mutate(y = diagnosed / registered) %>%
  custom_boxplot() +
  labs(
    y = "Proportion of TB among registered patients",
    title = "B | Prevalence of TB patients",
  ) +
  scale_y_continuous(
    limits = c(0, .5),
    expand = expansion(add = c(.02, 0)),
    labels = function(x) x * 100
  )
```

```{r modelled-risk}
well_riley <- df %>%
  dplyr::select(year, date, daytime, diagnosed, aer) %>%
  rename(I = diagnosed) %>%
  mutate(
    q = 2,
    t = 1,
    p = 0.5,
    V = 174.075,
    Q = aer * V,
    P = 1 - exp(-I * p * q * t / Q)
  )

well_riley_pl <- well_riley %>%
  rename(y = P) %>%
  custom_boxplot() +
  labs(
    y = "Probability of infection (%)",
    title = "E | Modelled transmission risk",
  ) +
  scale_y_continuous(
    limits = c(0, .05),
    expand = expansion(add = c(.001, 0)),
    labels = function(x) x * 100
  )

well_riley_pl
```

```{r molecular}
mol_pl <- df %>%
  filter(!is.na(dna_copies)) %>%
  mutate(
    y = log1p(dna_copies / sampling_time)
  ) %>%
  custom_boxplot() +
  labs(
    y = expression("Log DNA copies per " * mu * "L per hour of sampling"),
    title = "F | Mtb concentration in bioaerosols",
  ) +
  scale_y_continuous(
    limits = c(0, 10),
    expand = expansion(add = c(0.5, 0))
  )
```

**Figure**: Comparison of Mtb bioaerosol concentration, clinical occupancy, prevalence of TB patients, and environmental conditions between 2019 and 2021.

```{r summary-plot, fig.height=7.9, fig.width=6.3}
descriptive_pl <- track_pl + inc_pl + co2_pl + aer_pl + well_riley_pl + mol_pl +
  plot_layout(ncol = 2, nrow = 3, guides = "collect") &
  theme(legend.position = "bottom")

save_plot(
  descriptive_pl,
  "results/descriptives.png",
  w = 16,
  h = 20
)
```

Interpretation:

- Bioaerosol concentration was similar.
- Higher prevalence of TB patients among registered patients in 2019, which we would expect to reduce the outcome.
- Person-time in 2021 was much lower, which should reduce the outcome, unless the person-time of TB patients was similar (which we don't know).
- Air exchange rate was higher in 2021, suggesting better ventilation, which should reduce outcome.
- Relative humidity was lower and temperature higher in 2021, which could impact outcome, but the influence is unclear.

**Table**: Daytime-adjusted p-values from linear modeles to test for an overall difference in each variable between study periods.

```{r summary-pvalue}
p_value_overall <- function(dat, fam = "gaussian") {
  my <- glm(y ~ year + daytime, data = dat, family = fam)
  pval <- summary(my)$coefficients["year2021", 4]
  p_value_format(pval)
}

pvals_overall <- data.frame(
  variable = c(
    "Person-time in clinic",
    "CO2 concentration",
    "Air exchange rate",
    "Temperature",
    "Relative humidity",
    "Diagnosed TB",
    "Registed patients",
    "Prevalence of TB patients",
    "Modelled transmission risk",
    "Mtb concentration in bioaerosols"
  ),
  pvalue = c(
    p_value_overall(pt_df),
    p_value_overall(daily_max_co2 %>% rename(y = co2)),
    p_value_overall(df %>% rename(y = aer)),
    p_value_overall(df %>% rename(y = temperature)),
    p_value_overall(df %>% rename(y = humidity)),
    p_value_overall(df %>% mutate(y = diagnosed / registered)),
    p_value_overall(df %>% mutate(y = diagnosed), fam = "poisson"),
    p_value_overall(df %>% mutate(y = registered), fam = "poisson"),
    p_value_overall(well_riley %>% rename(y = P)),
    p_value_overall(df %>% mutate(y = log1p(dna_copies / sampling_time)))
  )
)

knitr::kable(pvals_overall, align = "rl")
```


### Comparison of TB patient characteristics

**Figure**: Comparison of the study periods in 2019 and 2021 regarding the proportion of positive GeneXpert test results, smear grading in TB patients, and proportion of HIV+ TB patients.

```{r clinical-characteristics, fig.height=4.7, fig.width=7}
col_pal <- paletteer_d("colorBlindness::Blue2Orange12Steps")
no_yes_idx <- c(4, 8)
x_exp <- 0

genexpert_pl <- clin %>%
  dplyr::select(year, xpert_pos, xpert_neg) %>%
  rename(
    `Pos` = xpert_pos,
    `Neg` = xpert_neg
  ) %>%
  reshape2::melt("year") %>%
  mutate(variable = factor(variable, levels = c("Neg", "Pos"))) %>%
  group_by(year) %>%
  mutate(
    prop = round(100 * value / sum(value))
  ) %>%
  ungroup() %>%
  mutate(lab = paste0(variable, ": ", value, " (", prop, "%)")) %>%
  ggplot(aes(x = prop, y = year, fill = variable)) +
  geom_bar(position = "stack", stat = "identity") +
  geom_text(
    aes(label = lab),
    size = 8 / cm(1),
    position = position_stack(vjust = 0.6)
  ) +
  scale_x_continuous(expand = expansion(add = c(0, x_exp))) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_fill_manual(values = col_pal[no_yes_idx]) +
  ggtitle("A | Proportion of positive GeneXpert test results among suspected patients") +
  theme_custom() +
  theme(
    axis.title = element_blank(),
    legend.position = "none"
  )

smear_pl <- clin %>%
  dplyr::select(year, matches("mear"), Negative) %>%
  reshape2::melt("year") %>%
  mutate(
    variable = gsub("Smear / Positive$", "Pos", variable),
    variable = gsub("Smear / Positive", "P", variable),
    variable = gsub("Smear / Scanty", "Scant", variable),
    variable = gsub("Negative", "Neg", variable),
  ) %>%
  filter(variable != "No smear") %>%
  mutate(
    variable = factor(
      variable,
      levels = c("Neg", "Scant", "Pos", "P+", "P++", "P+++")
    )
  ) %>%
  group_by(year) %>%
  mutate(
    prop = smart.round(100 * value / sum(value))
  ) %>%
  filter(prop > 0) %>%
  ungroup() %>%
  mutate(lab = paste0(variable, ": ", value, " (", prop, "%)")) %>%
  ggplot(aes(x = prop, y = year, fill = variable)) +
  geom_bar(position = "stack", stat = "identity") +
  geom_text(
    aes(label = lab),
    size = 8 / cm(1),
    position = position_stack(vjust = 0.375)
  ) +
  scale_x_continuous(
    expand = expansion(add = c(0, x_exp))
  ) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_fill_manual(values = col_pal[c(4, 8, 9, 10, 11)]) +
  ggtitle("B | Proportion of sputum smear gradings among TB patients") +
  theme_custom() +
  theme(
    axis.title = element_blank(),
    legend.position = "none"
  )

hiv_pl <- clin %>%
  dplyr::select(year, hiv_pos, hiv_neg) %>%
  rename(
    `HIV+` = hiv_pos,
    `HIV-` = hiv_neg
  ) %>%
  reshape2::melt("year") %>%
  mutate(variable = factor(variable, levels = c("HIV-", "HIV+"))) %>%
  group_by(year) %>%
  mutate(
    prop = round(100 * value / sum(value))
  ) %>%
  ungroup() %>%
  mutate(lab = paste0(variable, ": ", value, " (", prop, "%)")) %>%
  ggplot(aes(x = prop, y = year, fill = variable)) +
  geom_bar(position = "stack", stat = "identity") +
  geom_text(
    aes(label = lab),
    size = 8 / cm(1),
    position = position_stack(vjust = 0.5)
  ) +
  scale_x_continuous(expand = expansion(add = c(0, x_exp))) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_fill_manual(values = col_pal[no_yes_idx]) +
  ggtitle("C | Proportion of HIV-positive TB patients") +
  theme_custom() +
  theme(
    axis.title = element_blank(),
    legend.position = "none"
  )

clin_pl <- genexpert_pl + smear_pl + hiv_pl +
  plot_layout(ncol = 1, guides = "collect") &
  theme(legend.position = "none")

save_plot(
  clin_pl,
  "results/descriptives-clinical.png",
  w = 16,
  h = 12
)
```


## Supplementary results

**Figure**: Comparison of temperature and relative humidity between 2019 and 2021.

```{r supp-environ-temp-rH, fig.height=3, fig.width=6.3}
rH_pl <- df %>%
  filter(!is.na(humidity)) %>%
  rename(y = humidity) %>%
  custom_boxplot() +
  labs(
    y = "Relative humidity (%)",
    title = "A | Relative humidity",
  ) +
  scale_y_continuous(
    limits = c(0, 100),
    expand = c(0, 0)
  )

temp_pl <- df %>%
  filter(!is.na(temperature)) %>%
  rename(y = temperature) %>%
  custom_boxplot() +
  labs(
    y = "Temperature (°C)",
    title = "B | Temperature",
  ) +
  scale_y_continuous(
    limits = c(0, 35),
    expand = c(0, 0)
  )

supp_environ_descriptive_pl <- rH_pl + temp_pl +
  plot_layout(ncol = 2, nrow = 1, guides = "collect") &
  theme(legend.position = "bottom")

save_plot(
  supp_environ_descriptive_pl,
  "results/supp-environ-descriptives.png",
  w = 16,
  h = 12
)
```

&nbsp;

**Figure**: Coefficients of variation (standard deviation divided by mean).

```{r supp-variation}
recode_vars <- c(
  "pt" = "Person-time",
  "co2" = "CO2 concentration",
  "aer" = "Air exchange rate",
  "incidence" = "TB prevalence",
  "P" = "Modelled transmission risk",
  "outcome" = "Mtb concentration"
)

daily_df <- df %>%
  group_by(year, date) %>%
  filter(sum(!is.na(dna_copies)) == 2) %>%
  summarise(
    outcome = log1p((sum(dna_copies) / sum(sampling_time)))
  ) %>%
  ungroup() %>%
  left_join(
    df %>%
      group_by(year, date) %>%
      summarise(incidence = sum(diagnosed) / sum(registered))
  ) %>%
  left_join(
    df %>%
      group_by(year, date) %>%
      filter(sum(!is.na(aer)) == 2) %>%
      summarise(aer = mean(aer))
  ) %>%
  left_join(
    pt_df %>%
      rename(pt = y) %>%
      group_by(year, date) %>%
      filter(sum(!is.na(pt)) == 2) %>%
      summarise(pt = mean(pt))
  ) %>%
  left_join(
    daily_max_co2 %>%
      group_by(year, date) %>%
      filter(sum(!is.na(co2)) == 2) %>%
      summarise(co2 = max(co2))
  ) %>%
  left_join(
    well_riley %>%
      group_by(year, date) %>%
      filter(sum(!is.na(P)) == 2) %>%
      summarise(P = 1 - prod(1 - P))
  )

cv_df <- daily_df %>%
  mutate(outcome = expm1(outcome)) %>%
  reshape2::melt(c("year", "date")) %>%
  mutate(
    variable = recode(variable, !!!recode_vars),
    variable = factor(variable, levels = recode_vars)
  ) %>%
  group_by(year, variable) %>%
  summarise(
    mean = mean(value, na.rm = TRUE),
    median = median(value, na.rm = TRUE),
    sd = sd(value, na.rm = TRUE),
    q25 = quantile(value, .25, na.rm = TRUE),
    q75 = quantile(value, .75, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(
    cv = sd / mean,
    cv2 = (q75 - q25) / median
  )

cv_pl <- cv_df %>%
  ggplot(aes(x = year, fill = year, y = cv)) +
  facet_wrap(~variable, scales = "free_x") +
  geom_bar(
    stat = "identity",
    position = position_dodge(width = .75),
    width = .66
  ) +
  geom_text(
    aes(label = round(cv, 2)),
    size = 8 / cm(1),
    vjust = -1
  ) +
  scale_fill_manual(
    values = wes_palette("Zissou1")[c(2, 4)]
  ) +
  scale_y_continuous(
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.1))
  ) +
  theme_custom() +
  labs(y = "Coefficient of variation (SD / Mean)") +
  theme(
    strip.text.x = element_text(margin = margin(t = 12, b = 6)),
    legend.title = element_blank(),
    legend.position = "bottom",
    legend.key.width = unit(0.25, "cm"),
    legend.key.height = unit(0.3, "cm"),
    axis.title.x = element_blank()
  )

save_plot(
  cv_pl,
  pdf_file = "results/supp-coef-of-variation.png",
  w = 16, h = 16
)
```

Here the values as a table:

```{r supp-variation-tbl}
cv_df %>%
  dplyr::select(year, variable, cv2) %>%
  spread(year, cv2) %>%
  mutate(across(-variable, round_k, 2)) %>%
  knitr::kable(align = "lrr")
```

&nbsp;

In the following, we show the comparison of variables overall and by daytime. 
For the overall comparison of daily values, person-time, TB prevalence, and Mtb concentration were aggregated using the sum,
while environmental variables were aggregated using the mean of the morning and afternoon values;
for each variable, days with only morning or afternoon values were excluded.
In addition, CO2, temperature, and humidity are summarised using all values between 8am and 4pm. 

&nbsp;

```{r supp-functions}
summarise_iqr <- function(df, n_digits = 1) {
  df %>%
    summarise(
      median = median(value, na.rm = TRUE),
      lower = quantile(value, 0.25, na.rm = TRUE),
      upper = quantile(value, 0.75, na.rm = TRUE)
    ) %>%
    ungroup() %>%
    mutate_if(is.numeric, ~ format(round(.x, n_digits), nsmall = n_digits)) %>%
    mutate(value = paste0(median, " (", lower, ", ", upper, ")"))
}
```

&nbsp;

**Table**: Mtb concentration in bioaerosols by study year (median and IQR).

```{r mol-supp}
df %>%
  filter(!is.na(dna_copies)) %>%
  group_by(year, date) %>%
  filter(n() == 2) %>%
  mutate(
    value = log1p(sum(dna_copies) / sum(sampling_time))
  ) %>%
  ungroup() %>%
  dplyr::select(year, value) %>%
  group_by(year) %>%
  summarise_iqr() %>%
  ungroup() %>%
  dplyr::select(year, value) %>%
  arrange(year) %>%
  pivot_wider(names_from = c(year), values_from = value) %>%
  knitr::kable(align = "lrr")
```

&nbsp;

**Table**: Mtb concentration in bioaerosols by study year and daytime (median and IQR).

```{r mol-supp-daytime}
df %>%
  filter(!is.na(dna_copies)) %>%
  mutate(
    value = log1p(dna_copies / sampling_time)
  ) %>%
  dplyr::select(year, daytime, value) %>%
  group_by(year, daytime) %>%
  summarise_iqr() %>%
  ungroup() %>%
  dplyr::select(year, daytime, value) %>%
  arrange(year, desc(daytime)) %>%
  pivot_wider(names_from = c(year, daytime), values_from = value) %>%
  knitr::kable(align = "lrrrr")
```

&nbsp;

**Table**: Modelled risk of transmission by study year (median and IQR).

```{r mol-supp-risk}
well_riley %>%
  group_by(year, date) %>%
  filter(sum(!is.na(P)) == 2) %>%
  summarise(P = 1 - prod(1 - P)) %>%
  ungroup() %>%
  mutate(value = P * 100) %>%
  dplyr::select(year, value) %>%
  group_by(year) %>%
  summarise_iqr(n_digits = 2) %>%
  ungroup() %>%
  dplyr::select(year, value) %>%
  arrange(year) %>%
  pivot_wider(names_from = c(year), values_from = value) %>%
  knitr::kable(align = "lrr")
```

&nbsp;

**Table**: Modelled risk of transmission by study year and daytime (median and IQR).

```{r mol-supp-risk-daytime}
well_riley %>%
  mutate(value = P * 100) %>%
  dplyr::select(year, daytime, value) %>%
  group_by(year, daytime) %>%
  summarise_iqr(n_digits = 2) %>%
  ungroup() %>%
  dplyr::select(year, daytime, value) %>%
  arrange(year, desc(daytime)) %>%
  pivot_wider(names_from = c(year, daytime), values_from = value) %>%
  knitr::kable(align = "lrrrr")
```

&nbsp;

**Table**: Environmental conditions by study year (median and IQR).

```{r environ-supp}
df %>%
  left_join(daily_max_co2) %>%
  dplyr::select(year, date, co2, aer, humidity, temperature) %>%
  reshape2::melt(c("year", "date")) %>%
  group_by(year, date, variable) %>%
  filter(sum(!is.na(value)) == 2) %>%
  ungroup() %>%
  group_by(year, variable) %>%
  summarise_iqr(n_digits = 0) %>%
  dplyr::select(year, variable, value) %>%
  arrange(year) %>%
  pivot_wider(names_from = c(year), values_from = value) %>%
  knitr::kable(align = "lrr")
```

&nbsp;

**Table**: Environmental conditions by study year and daytime (median and IQR).

```{r environ-supp-daytime}
df %>%
  left_join(daily_max_co2) %>%
  dplyr::select(year, daytime, co2, aer, humidity, temperature) %>%
  reshape2::melt(c("year", "daytime")) %>%
  group_by(year, daytime, variable) %>%
  summarise_iqr(n_digits = 0) %>%
  dplyr::select(year, daytime, variable, value) %>%
  arrange(year, desc(daytime)) %>%
  pivot_wider(names_from = c(year, daytime), values_from = value) %>%
  knitr::kable(align = "lrrrr")
```

&nbsp;

**Table**: Person-time in the clinic (hours) by study year (median and IQR). 

```{r tracking-supp}
pt_df %>%
  rename(value = y) %>%
  group_by(year, date) %>%
  filter(sum(!is.na(value)) == 2) %>%
  summarise(value = sum(value)) %>%
  ungroup() %>%
  group_by(year) %>%
  summarise_iqr(n_digits = 0) %>%
  dplyr::select(year, value) %>%
  arrange(year) %>%
  pivot_wider(names_from = c(year), values_from = value) %>%
  knitr::kable(align = "lrr")
```

&nbsp;

**Table**: Person-time in the clinic (hours) by study year and daytime (median and IQR).

```{r tracking-supp-daytime}
pt_df %>%
  rename(value = y) %>%
  group_by(year, daytime) %>%
  summarise_iqr(n_digits = 0) %>%
  dplyr::select(year, daytime, value) %>%
  arrange(year, desc(daytime)) %>%
  pivot_wider(names_from = c(year, daytime), values_from = value) %>%
  knitr::kable(align = "lrrrr")
```

&nbsp;

**Table**: Prevalence (diagnosed/registered in %) and number of registered and diagnosed TB patients by study year (median and IQR).

```{r tb-visits-supp}
df %>%
  mutate(prevalence = diagnosed / registered * 100) %>%
  dplyr::select(year, date, prevalence, registered, diagnosed) %>%
  reshape2::melt(c("year", "date")) %>%
  group_by(year, date, variable) %>%
  summarise(value = sum(value)) %>%
  ungroup() %>%
  group_by(year, variable) %>%
  summarise_iqr(n_digits = 0) %>%
  dplyr::select(year, variable, value) %>%
  pivot_wider(names_from = c(year), values_from = value) %>%
  knitr::kable(align = "lrr")
```

&nbsp;

**Table**: Prevalence (diagnosed/registered in %) and number of registered and diagnosed TB patients by study year and daytime (median and IQR).

```{r tb-visits-supp-daytime}
df %>%
  mutate(prevalence = diagnosed / registered * 100) %>%
  dplyr::select(year, daytime, prevalence, registered, diagnosed) %>%
  reshape2::melt(c("year", "daytime")) %>%
  group_by(year, daytime, variable) %>%
  summarise_iqr(n_digits = 0) %>%
  dplyr::select(year, daytime, variable, value) %>%
  pivot_wider(names_from = c(year, daytime), values_from = value) %>%
  knitr::kable(align = "lrrrr")
```


&nbsp;

**Table**: Clinical data by study years (counts).

```{r clinical-characteristics-supp}
clin %>%
  left_join(
    df %>%
      group_by(year) %>%
      summarise(days = n_distinct(date)) %>%
      ungroup()
  ) %>%
  mutate(
    tbr = cases / days,
    csr = tb_patients / suspected * 100
  ) %>%
  rename(
    `Number of study days` = days,
    `Diagnosed TB patients` = tb_patients,
    `Suspected TB cases` = suspected,
    `Multi-drugresistant TB cases` = mdr,
    `Relapse TB cases` = relapses,
    `HIV positive` = hiv_pos,
    `HIV negative` = hiv_neg,
    `Smear negative result` = Negative,
    `Smear positive result` = Positive,
    `GeneXpert positive result` = xpert_pos,
    `GeneXpert negative result` = xpert_neg
  ) %>%
  dplyr::select(
    year,
    `Number of study days`,
    `Diagnosed TB patients`,
    `Suspected TB cases`,
    `Multi-drugresistant TB cases`,
    `Relapse TB cases`,
    `HIV positive`,
    `HIV negative`,
    `Smear positive result`,
    `Smear / Positive`,
    `Smear / Positive+`,
    `Smear / Positive++`,
    `Smear / Positive+++`,
    `Smear / Scanty`,
    `Smear negative result`,
    `GeneXpert positive result`,
    `GeneXpert negative result`
  ) %>%
  reshape2::melt("year") %>%
  pivot_wider(names_from = year, values_from = value) %>%
  knitr::kable(align = "lrr")
```

&nbsp;

**Figure**: Pairwise correlation between bioaerosol concentration, person-time in clinic, air exchange rate, relative humidity, and temperature.

```{r correlation, fig.height=6.3}
cor_pl <- df %>%
  mutate(log_dna_hr = log1p(dna_copies / sampling_time)) %>%
  left_join(pt_df, by = c("year", "date", "daytime")) %>%
  left_join(daily_max_co2, , by = c("year", "date", "daytime")) %>%
  rename(person_time = y) %>%
  dplyr::select(
    year, daytime,
    log_dna_hr,
    person_time,
    co2, aer, humidity, temperature
  ) %>%
  ggpairs(
    columns = 3:ncol(.),
    mapping = aes(color = year, fill = year, alpha = .5),
    columnLabels = c(
      "Conc. (log/h)",
      "Person-time (h)",
      "CO2 (ppm)",
      "AER (1/h)",
      "RH (%)",
      "Temperature (°C)"
    ),
    upper = list(
      continuous = GGally::wrap(
        ggally_cor,
        use = "pairwise.complete.obs",
        size = 8 / cm(1),
        title = "r",
        digits = 2
      )
    ),
  ) +
  scale_fill_manual(
    values = wes_palette("Zissou1")[c(2, 4)]
  ) +
  scale_color_manual(
    values = wes_palette("Zissou1")[c(2, 4)]
  ) +
  theme_custom() +
  theme(
    strip.text = element_text(margin = margin(b = 10, l = 10)),
    panel.spacing = unit(0.5, "cm")
  )

save_plot(
  cor_pl,
  "results/correlation.png",
  w = 16,
  h = 16
)
```

Interpretation:

- Mtb concentration not correlated with person time in the clinic or environmental conditions.

&nbsp;

**Figure**: Comparison of the entire years 2019 and 2021 regarding the proportion of positive GeneXpert test results, smear grading in TB patients, and proportion of HIV+ TB patients.

```{r clin-data-full-years, include=FALSE}
clin_full <- read_csv("data-clean/clinical-data-full-years.csv") %>%
  mutate(year = factor(year, levels = c(2019, 2021)))
```

```{r clinical-characteristics-full-years, fig.height=4.7, fig.width=7}
genexpert_full_pl <- clin_full %>%
  dplyr::select(year, xpert_pos, xpert_neg) %>%
  rename(
    `Pos` = xpert_pos,
    `Neg` = xpert_neg
  ) %>%
  reshape2::melt("year") %>%
  mutate(variable = factor(variable, levels = c("Neg", "Pos"))) %>%
  group_by(year) %>%
  mutate(
    prop = round(100 * value / sum(value))
  ) %>%
  ungroup() %>%
  mutate(lab = paste0(variable, ": ", value, " (", prop, "%)")) %>%
  ggplot(aes(x = prop, y = year, fill = variable)) +
  geom_bar(position = "stack", stat = "identity") +
  geom_text(
    aes(label = lab),
    size = 8 / cm(1),
    position = position_stack(vjust = 0.6)
  ) +
  scale_x_continuous(expand = expansion(add = c(0, x_exp))) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_fill_manual(values = col_pal[no_yes_idx]) +
  ggtitle("A | Proportion of positive GeneXpert test results among suspected patients") +
  theme_custom() +
  theme(
    axis.title = element_blank(),
    legend.position = "none"
  )

smear_full_pl <- clin_full %>%
  dplyr::select(year, matches("mear"), Negative) %>%
  reshape2::melt("year") %>%
  mutate(
    variable = gsub("Smear / Positive$", "Pos", variable),
    variable = gsub("Smear / Positive", "P", variable),
    variable = gsub("Smear / Scanty", "Scant", variable),
    variable = gsub("Negative", "Neg", variable),
  ) %>%
  filter(variable != "No smear") %>%
  mutate(
    variable = factor(
      variable,
      levels = c("Neg", "Scant", "Pos", "P+", "P++", "P+++")
    )
  ) %>%
  group_by(year) %>%
  mutate(
    prop = smart.round(100 * value / sum(value))
  ) %>%
  filter(prop > 0) %>%
  ungroup() %>%
  mutate(lab = paste0(variable, ": ", value, " (", prop, "%)")) %>%
  ggplot(aes(x = prop, y = year, fill = variable)) +
  geom_bar(position = "stack", stat = "identity") +
  geom_text(
    aes(label = lab),
    size = 8 / cm(1),
    position = position_stack(vjust = 0.375)
  ) +
  scale_x_continuous(
    expand = expansion(add = c(0, x_exp))
  ) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_fill_manual(values = col_pal[c(4, 8, 9, 10, 11)]) +
  ggtitle("B | Proportion of sputum smear gradings among TB patients") +
  theme_custom() +
  theme(
    axis.title = element_blank(),
    legend.position = "none"
  )

hiv_full_pl <- clin_full %>%
  dplyr::select(year, hiv_pos, hiv_neg) %>%
  rename(
    `HIV+` = hiv_pos,
    `HIV-` = hiv_neg
  ) %>%
  reshape2::melt("year") %>%
  mutate(variable = factor(variable, levels = c("HIV-", "HIV+"))) %>%
  group_by(year) %>%
  mutate(
    prop = round(100 * value / sum(value))
  ) %>%
  ungroup() %>%
  mutate(lab = paste0(variable, ": ", value, " (", prop, "%)")) %>%
  ggplot(aes(x = prop, y = year, fill = variable)) +
  geom_bar(position = "stack", stat = "identity") +
  geom_text(
    aes(label = lab),
    size = 8 / cm(1),
    position = position_stack(vjust = 0.5)
  ) +
  scale_x_continuous(expand = expansion(add = c(0, x_exp))) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_fill_manual(values = col_pal[no_yes_idx]) +
  ggtitle("C | Proportion of HIV-positive TB patients") +
  theme_custom() +
  theme(
    axis.title = element_blank(),
    legend.position = "none"
  )

clin_full_pl <- genexpert_full_pl + smear_full_pl + hiv_full_pl +
  plot_layout(ncol = 1, guides = "collect") &
  theme(legend.position = "none")

save_plot(
  clin_full_pl,
  "results/descriptives-clinical-full.png",
  w = 16,
  h = 12
)
```